% IEEE Double-Column Template with Appendix
% Compile: pdflatex → bibtex → pdflatex → pdflatex

\documentclass[conference]{IEEEtran}

\IEEEoverridecommandlockouts

\usepackage[numbers]{natbib}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{url}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{listings}

% Code listing style
\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  captionpos=b
}

\begin{document}

% ---------- TITLE ----------
\title{Intelligent Learning Analytics and Agentic Study Coach}

\author{
\IEEEauthorblockN{Team Members Name separated by Comma}
\IEEEauthorblockA{
Team Name \\}
}

\maketitle

% ---------- ABSTRACT ----------
\begin{abstract}
This paper presents an end-to-end Intelligent Learning Analytics and Agentic Study Coach system. In Milestone 1, we built a machine learning pipeline using Logistic Regression to classify students into At-risk, Average, and High-performing categories with approximately 70\% test accuracy, and generated baseline rule-based recommendations. In Milestone 2, we extended the system into an agentic study coach that accepts student goals, diagnoses learning gaps, retrieves relevant learning resources using a free arXiv API workflow, and generates structured multi-week study plans through an open-source/free-tier LLM pipeline with robust fallback logic. The final system is deployed through a Streamlit interface with downloadable Markdown/PDF reports and session-based memory for iterative coaching.
\end{abstract}

\begin{IEEEkeywords}
Educational data mining, learning analytics, Logistic Regression, agentic AI, study planning, retrieval-augmented recommendations.
\end{IEEEkeywords}

% ---------- INTRODUCTION ----------
\section{Introduction}
The educational landscape is increasingly utilizing data to improve student outcomes. Identifying students who may be at risk of academic failure early on allows for timely intervention. This project aims to bridge the gap between performance data and actionable study plans.

% ---------- RELATED WORK ----------
\section{Related Work}
Previous research in Educational Data Mining (EDM) has explored various classification techniques, including Decision Trees, SVMs, and Neural Networks, to predict student success based on demographic and historical grade data.

% ---------- METHODOLOGY ----------
\section{Methodology}
The methodology is organized in two layers: (1) ML-based analytics for performance prediction and (2) an agentic coaching layer for personalized planning.

\subsection{System Workflow}
The final application follows a hybrid analytics + agentic pipeline:
\begin{enumerate}
    \item \textbf{Data Ingestion}: Raw student performance data (CSV) is uploaded via the dashboard.
    \item \textbf{Preprocessing and Feature Engineering}: Categorical demographic features are label-encoded; total and percentage scores are computed to derive performance categories.
    \item \textbf{ML Inference (Milestone 1)}: A trained Logistic Regression model predicts student risk/performance labels.
    \item \textbf{Goal Capture (Milestone 2)}: The student provides a learning goal, study style, available weekly time, and study horizon.
    \item \textbf{Learning Diagnosis}: The agentic layer identifies strengths, gap areas, and risk level using student scores and derived indicators.
    \item \textbf{Resource Retrieval}: Gap-aware queries are sent to the arXiv API to collect relevant learning resources and URLs.
    \item \textbf{Plan Generation}: An LLM-based module creates a structured report with diagnosis, personalized plan, weekly goals, resources, and next steps.
    \item \textbf{Fallback and Delivery}: If LLM inference is unavailable, a rule-based planner generates the same required structured output; reports are shown in UI and exported as Markdown/PDF.
\end{enumerate}

\subsection{Dataset Description}
The dataset consists of 1,000 student records from a simulated exam environment, featuring eight attributes: gender, race/ethnicity, parental education, lunch type, test preparation course, math score, reading score, and writing score.

\subsection{Preprocessing}
Categorical variables were encoded using Label Encoding. Scores were aggregated to calculate total percentages. The data was scaled using Standard Scaler before training models.

\subsection{Feature Engineering}
A new feature, "performance category," was derived from the average of the three test scores, serving as the target variable for multi-class classification.

\subsection{Models Used}
We used two model components:
\begin{itemize}
    \item \textbf{Logistic Regression} for supervised student performance classification (Milestone 1).
    \item \textbf{Open-source/free-tier LLM workflow} for plan generation (Milestone 2), using Hugging Face inference when token access is available, otherwise local open-source inference, and a deterministic fallback template if model calls fail.
\end{itemize}

\subsection{Training and Validation}
The dataset was split into 80\% training and 20\% testing sets. Performance was evaluated using Accuracy, Precision, Recall, and F1-score.

% ---------- RESULTS ----------
\section{Results}

\subsection{Quantitative Results}
The Logistic Regression model was trained to predict student performance levels based on demographic data. The overall accuracy achieved was 70\%. Table I summarizes the performance across different metrics.

\begin{table}[ht]
\centering
\caption{Model Performance Metrics}
\begin{tabular}{lcccc}
\toprule
Model & Acc & Prec & Rec & F1 \\
\midrule
Logistic Regression & 0.70 & 0.54 & 0.70 & 0.59 \\
\bottomrule
\end{tabular}
\end{table}

To provide dataset-level context for the model performance, Table II reports key descriptive statistics and class distribution after deriving performance categories from average score thresholds.

\begin{table}[ht]
\centering
\caption{Dataset Summary (StudentsPerformance.csv)}
\begin{tabular}{lc}
\toprule
Statistic & Value \\
\midrule
Total students & 1000 \\
Mean math score & 66.09 \\
Mean reading score & 69.17 \\
Mean writing score & 68.05 \\
Mean overall score & 67.77 \\
At-risk count & 103 \\
Average count & 699 \\
High-performing count & 198 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Milestone 2 Functional Results}
The deployed agentic coach successfully generates the required structured report sections for user-defined goals:
\begin{itemize}
    \item Learning diagnosis (risk, strengths, and learning gaps),
    \item Personalized multi-week study plan,
    \item Weekly milestones/goals,
    \item Recommended resources with URLs,
    \item Progress feedback and next-step guidance.
\end{itemize}
The system supports session-level coaching history and report export (Markdown/PDF). A robustness fix was integrated in the PDF pipeline to safely handle long unbroken tokens (such as URLs), preventing rendering failures in deployment environments.

\subsection{Qualitative Observations}
Exploratory analysis indicated that math, reading, and writing scores are positively related, and distribution-level behavior was consistent with the class balance shown in Table II.

% ---------- DISCUSSION ----------
\section{Discussion}
While the baseline classifier offers useful risk stratification, precision for identifying At-risk students remains limited when relying mainly on demographic and exam aggregates. The agentic extension improves practical usefulness by converting model outputs and score gaps into actionable, time-aware study plans. Resource retrieval from arXiv provides transparent, citeable references, but retrieved materials can be research-oriented for some learners. LLM-based plan quality depends on model availability and latency; therefore, a deterministic fallback planner was included to preserve reliability and maintain required output structure under constrained free-tier conditions.

% ---------- CONCLUSION ----------
\section{Conclusion}
This work demonstrates a complete progression from classical learning analytics to an agentic AI study coach. Logistic Regression provided an interpretable baseline for student risk classification, and the Milestone 2 extension operationalized these insights through goal-driven diagnosis, resource retrieval, and personalized study planning in a publicly deployable interface. The resulting system satisfies the required educational outputs while remaining compatible with free/open-source tooling. Future work can include persistent longitudinal memory, adaptive difficulty calibration, and domain-specific retrieval corpora to improve pedagogical alignment.

% ---------- REFERENCES ----------
\bibliographystyle{IEEEtran}
\bibliography{references}

% ---------- APPENDIX ----------
\appendix

\section{GitHub Repository Information}

This appendix provides details regarding the project’s GitHub repository and associated file structure for reproducibility and open-source collaboration.

\subsection{Repository Link}
The code and docs are available at:

\begin{center}
\textbf{\url{https://github.com/LuciferVid/learning_analytics}}
\end{center}

\subsection{Repository Structure}

The project follows this layout:
\begin{lstlisting}
.
|-- data/ (datasets)
|-- notebooks/ (EDA)
|-- src/ (preprocessing, models, coach modules)
|-- models/ (trained pkls)
|-- results/ (plots)
|-- app.py (streamlit app)
|-- main.tex (this report)
`-- requirements.txt
\end{lstlisting}

\subsection{Key Components}

\begin{itemize}
    \item \textbf{data/}: Raw student score data.
    \item \textbf{notebooks/}: Initial analysis and experiments.
    \item \textbf{src/}: Python modules for preprocessing, ML inference, agentic coaching, retrieval, and PDF export.
    \item \textbf{models/}: Saved model, scaler, and encoders.
    \item \textbf{results/}: Generated graphs and tables.
    \item \textbf{app.py}: Interactive Streamlit dashboard with Milestone 1 and Milestone 2 workflows.
\end{itemize}

\end{document}
